\section{Current Progress and Future Work}

Two parallel development branches have been pursued in this project. Each branch implements the same core framework described in the previous sections, namely a residual RL policy on top of a mesoscopic adaptive headway rule, but the two branches differ in their engineering choices and experimental focus. In this section, we summarize the current state of each branch, compare the experimental results obtained so far, and outline the remaining work.

\subsection{Branch overview}

\paragraph{Branch~A (sequential CPU implementation).}
This branch follows a lightweight, single-episode training loop running entirely on CPU. The RL module is embedded inside the mesoscopic adaptation layer: the rule-based adapter first computes a headway scaling factor~$\alpha^{\mathrm{rule}}$, and the RL policy then outputs a bounded residual correction $\Delta\alpha \in [-0.1, 0.1]$. The observation is an 8-dimensional vector that captures both local car-following variables and upstream mesoscopic statistics:
\[
x_i = \bigl(\mu_v,\;\sigma_v^2,\;\delta_v,\;v_i,\;s_i,\;\Delta v_i,\;a_{\ell(i)},\;\alpha_i^{\mathrm{prev}}\bigr).
\]
Training collects one full episode (1500~time steps) before each policy update. Both PPO and SAC have been implemented and compared over 800~training episodes.

\paragraph{Branch~B (vectorized GPU implementation).}
This branch rewrites the entire ring-road simulator as a set of parallel PyTorch tensor operations, eliminating all Python loops over vehicles. Multiple independent environments (16 to 128) are batched on the GPU, so that a single \texttt{step()} call advances all simulations simultaneously. The RL layer is decoupled from the mesoscopic module and placed in its own \texttt{agents} package. The observation space is extended to 15~dimensions by adding global context features such as fleet mean speed, CAV penetration rate, and neighbourhood statistics. GPU-native PPO and SAC trainers have been implemented, together with a formal evaluation protocol that sweeps over four control modes (baseline, adaptive, PPO, SAC), four human-driver ratios, two CAV layout strategies (ordered vs.\ shuffled), and optional perturbation injection.

\subsection{Experimental results}

Both branches evaluate three control modes, no~RL (baseline adaptive headway), PPO, and SAC, across several human-driver penetration rates on the same 22-vehicle, 300\,m ring road. The key metrics are mean fleet speed, speed variance, minimum inter-vehicle gap, collision count, and ride comfort (RMS acceleration and jerk).

\subsubsection{Branch~A results (CPU, 800 episodes)}

Table~\ref{tab:branch_a_results} summarizes the evaluation results for Branch~A. Each row corresponds to one control mode evaluated at a given human-driver ratio. The most critical scenario is the 75\,\% human-driver case, where the traffic is dominated by stochastic HDV behavior.

\begin{table}[!htbp]
\centering
\caption{Branch~A evaluation metrics across human-driver ratios (800~episodes, CPU training).}
\label{tab:branch_a_results}
\small
\begin{tabular}{llcccccc}
\hline
HR & Mode & Mean speed & Speed var. & RMS acc. & RMS jerk & Min gap & Collisions \\
 & & (m/s) & (m$^2$/s$^2$) & (m/s$^2$) & (m/s$^3$) & (m) & \\
\hline
\multirow{3}{*}{25\,\%}
 & No RL  & 4.868 & 0.630 & 0.578 & 8.682 & 0.510 & 0 \\
 & PPO    & \textbf{5.081} & 0.780 & \textbf{0.441} & \textbf{6.302} & 0.510 & 0 \\
 & SAC    & 4.822 & \textbf{0.630} & 0.746 & 11.363 & 0.510 & 0 \\
\hline
\multirow{3}{*}{50\,\%}
 & No RL  & 3.745 & 1.939 & \textbf{0.437} & \textbf{5.169} & 0.510 & 0 \\
 & PPO    & 3.727 & \textbf{1.819} & 0.439 & 5.257 & 0.510 & 0 \\
 & SAC    & \textbf{3.755} & 1.916 & 0.477 & 6.011 & 0.510 & 0 \\
\hline
\multirow{3}{*}{75\,\%}
 & No RL  & 2.982 & 3.793 & 0.420 & \textbf{3.451} & 0.307 & 42 \\
 & PPO    & \textbf{2.992} & \textbf{3.608} & \textbf{0.419} & 3.520 & \textbf{0.489} & \textbf{1} \\
 & SAC    & 2.977 & 3.829 & 0.426 & 3.591 & 0.300 & 42 \\
\hline
\end{tabular}
\end{table}

At 25\,\% human drivers, all three methods perform well and no collisions occur. PPO achieves the highest mean speed (5.08\,m/s) and the best comfort metrics, while SAC matches the baseline variance. At 50\,\%, the three modes are comparable, with PPO slightly reducing speed variance.

The most significant result appears at 75\,\% human drivers. The baseline and SAC both produce 42~collisions, whereas PPO reduces this to a single collision, a 98\,\% improvement. PPO also achieves the best speed variance (3.608 vs.\ 3.793) and the largest minimum gap (0.489\,m vs.\ 0.307\,m). This suggests that the on-policy nature of PPO is better suited to the high-stochasticity regime of this scenario, where HDV behavior dominates.

\subsubsection{Branch~B results (GPU, PPO evaluation)}

Branch~B evaluates five training checkpoints across the same human-driver ratios. Additionally, it introduces a perturbation-injection protocol and a formal string-stability metric (the amplification ratio: the ratio of the maximum downstream speed perturbation to the upstream perturbation). Table~\ref{tab:branch_b_results} summarizes the results for the final PPO checkpoint compared with the adaptive and baseline modes.

\begin{table}[!htbp]
\centering
\caption{Branch~B evaluation metrics (GPU PPO final checkpoint, overall summary).}
\label{tab:branch_b_results}
\small
\begin{tabular}{lcccccc}
\hline
Mode & HR & Mean speed & Speed var. & Min gap & Safety rate & Amplification \\
 & & (m/s) & (m$^2$/s$^2$) & (m) & (\%) & ratio \\
\hline
\multirow{3}{*}{Adaptive}
 & 0\,\%  & 4.829 & 0.321 & 7.32 & 100 & 0.525 \\
 & 25\,\% & 4.975 & 0.011 & --- & 100 & --- \\
 & 50\,\% & 4.587 & 0.232 & --- & --- & --- \\
\hline
\multirow{3}{*}{Baseline}
 & 0\,\%  & 5.019 & 0.040 & 7.41 & 100 & 0.368 \\
 & 25\,\% & 4.975 & 0.011 & --- & 100 & --- \\
 & 50\,\% & 4.330 & 1.229 & --- & --- & --- \\
\hline
\multirow{3}{*}{RL (PPO)}
 & 0\,\%  & \textbf{7.081} & 0.094 & 6.91 & 100 & 0.744 \\
 & 25\,\% & \textbf{6.865} & 0.138 & --- & --- & --- \\
 & 50\,\% & 4.499 & 1.145 & --- & --- & --- \\
\hline
\end{tabular}
\end{table}

The GPU-trained PPO agent achieves a substantial speed improvement in the low human-driver regime: +46.5\,\% over adaptive at 0\,\% humans and +38.1\,\% at 25\,\% humans. However, this comes at the cost of reduced string stability (amplification ratio 0.744, closer to the instability threshold of~1.0). The checkpoint ranking analysis reveals an inherent speed--safety tradeoff: later checkpoints produce higher speeds but progressively worse safety metrics. The best safety--performance tradeoff is found at an early checkpoint (step~229\,376), which achieves a 50\,\% safety pass rate across all evaluated configurations.

\subsection{Comparison between the two branches}

Although the two branches share the same underlying framework, several design differences lead to complementary strengths and weaknesses.

\paragraph{Observation space.}
Branch~A uses 8~dimensions (local state only), while Branch~B uses 15~dimensions (local + global context). The richer observation may help the policy generalize across different penetration rates, but it also increases the number of hyperparameters and makes the learned policy harder to interpret.

\paragraph{RL--mesoscopic coupling.}
In Branch~A, the RL policy operates on top of the mesoscopic adapter, which remains active at runtime and provides a stable baseline. In Branch~B, the RL agent is decoupled from the mesoscopic layer. This means that Branch~A always benefits from the rule-based fallback, while Branch~B gives the RL agent more freedom but also more responsibility.

\paragraph{Training scale.}
Branch~A trains from single-episode rollouts on CPU (approximately 8~hours for 800~episodes). Branch~B parallelizes across 16--128~environments per GPU step, enabling over an order of magnitude more transitions per wall-clock second. This allows Branch~B to explore a much larger portion of the state space, which may explain the aggressive speed gains observed at low human-driver ratios.

\paragraph{Safety formalization.}
Branch~A monitors collisions and gap clamps as logged quantities. Branch~B introduces a formal safety-first checkpoint ranking: checkpoints are first filtered by worst-case collision count, then by minimum gap, and finally by traffic performance. This ensures that no checkpoint is selected purely on speed if it violates safety constraints.

\paragraph{String stability.}
Branch~A uses a binary validity criterion (no gap clamps during the episode). Branch~B computes a continuous amplification ratio from a controlled perturbation experiment. The continuous metric is more informative for evaluating the degree of string stability across conditions.

\subsection{Discussion}

The results so far show that both PPO and SAC can improve traffic behavior compared to the rule-based adaptive baseline, but their advantages depend on the traffic regime and the evaluation metric.

In the high human-driver regime (75\,\%), Branch~A's PPO achieves a dramatic safety improvement (42 collisions reduced to~1), suggesting that the coupled RL--mesoscopic design and the conservative observation space are well suited when HDV behavior dominates. In the low human-driver regime (0--25\,\%), Branch~B's GPU-trained PPO achieves much higher fleet speeds (+38--46\,\%), likely because the larger observation space and more extensive training allow the policy to exploit the coordinated nature of the all-CAV or mostly-CAV traffic.

An important finding from Branch~B is the speed--safety tradeoff across training checkpoints. As training progresses, the agent learns to achieve higher speeds but at the cost of smaller gaps and higher perturbation amplification. This suggests that future reward designs should include explicit safety constraints rather than relying on soft penalty terms alone.

SAC performs comparably to PPO in moderate traffic conditions (25--50\,\% human drivers) in both branches but struggles at 75\,\% in Branch~A. This may indicate that off-policy learning, while more sample-efficient in theory, is more sensitive to the distribution shift caused by highly stochastic HDV behavior in the single-episode training setting.

\subsection{Future work}

\paragraph{Unified evaluation.}
The most immediate next step is to evaluate both branches under an identical protocol: the same ring-road configuration, the same human-driver ratios, the same random seeds, and the same perturbation injections. This will allow a fair comparison between the two observation spaces, reward designs, and training scales.

\paragraph{Reward shaping and safety constraints.}
The speed--safety tradeoff observed in Branch~B motivates the introduction of constrained optimization techniques, such as Lagrangian relaxation or a safety critic, to enforce hard safety bounds during training rather than relying on post-hoc checkpoint filtering.

\paragraph{Merging the best of both branches.}
A natural synthesis would combine Branch~A's coupled RL--mesoscopic control stack (where the rule-based adapter remains active) with Branch~B's vectorized GPU simulator and richer observation space. This would retain the safety benefits of the layered architecture while enabling large-scale parallel training.

\paragraph{Extended traffic scenarios.}
The current evaluation is limited to a single ring-road configuration with 22~vehicles. Future experiments should vary the road length, vehicle count, and HDV aggressiveness distributions to assess how well the learned policies generalize beyond the training conditions.

\paragraph{SAC improvements.}
The poor collision performance of SAC at 75\,\% human drivers deserves further investigation. Possible directions include tuning the entropy temperature schedule, increasing the replay buffer size, or pre-training SAC from a PPO checkpoint (a capability already implemented in Branch~B).
